---
title: "Analyzing ATAC-seq samples with edgeCounter"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Analyzing ATAC-seq samples with edgeCounter}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

Assay for Transposase-Accessible Chromatin using Sequencing (ATAC-seq) is an 
alternative technique for assessing chromatin accessibility. In this method, 
hyperactive Tn5 transposase inserts transposon into the genome and transposition 
frequency depends mainly on 'genome accessibility'.

After aligning the ATAC-seq reads to the genome, the following tasks need to be 
performed for data interpretation:

1. Quality control assessment, which may be performed by `ATACseqQC` package
2. Counting transposition events given genomic ranges of interest
3. Perform differential analysis to understand how certain treatment affects 
genome accessibility as measurement by the transposition event counts.

While `ATACseqQC` package handles the quality control step satisfactorily, 
a well-documented and validated workflow is lacking for transposition event 
counting and downstream differential analysis.

`edgeCounter` fills this gap by providing functions for straightforward transposition 
counts which may be used for downstream differential analysis by packages such as 
`DESeq2`, `edgeR`, etc.

# Goal of the vignette

This vignette demonstrates a full workflow in which we:

1. Perform a subset of QC tests provided by the `ATACseqQC` package
2. Use `ATACseqQC` to shift the reads such that read edges (i.e., boundaries) 
represent true transposition location
3. Take the shifted reads and count transposition events given several example 
genomic ranges

For your own data, you need to first align and filter the ATAC-seq reads to get 
BAM files of the alignment output. We assume that BAM files are already available 
in this vignette.

# Sample dataset intro

The sample dataset is from a series of experiments in which ~1,000 neurons from 
Drosophila melanogaster brains are used for ATAC-seq. Only alignments close to 
the housekeeping Gapdh1 gene are included (`dm6 Chr2R:7791200-7793600`) totaling 
1194 reads (597 pairs) in order to reduce package size.

Given the small input, we expect less-than-ideal quality control metrics. 
However, key metrics such as read enrichment from transcription start sites should 
show expected results - see `ATACseqQC` package for more details.

**Important note:** To get a full picture of quality controls of your sample, you 
should use the full alignment file instead of a small subset. Therefore, in
Section '`ATACseqQC`: Quality controls and read shifting' we only document 
the subset of QC plots we preferred and do not show the output figures.


# Environment setup

```{r setup}
library(ATACseqQC)
library(edgeCounter)

sample.file <- system.file("extdata", "sample.near.Gapdh1.bam", 
                           package = "edgeCounter")
```

# `ATACseqQC`: Quality controls and read shifting

Initial quality controls such as fragment size distribution can be performed 
using the original BAM file. Quality controls relative to genome landmarks (e.g., 
TSS) is performed after shifting the alignments.

```{r ATACseqQC plots, echo=T, include=F}
# Fragment size distribution plot
fragSizeDist(sample.file, "Sample")

# Shift the reads to reflect true transposition site
#   might not be essential as shifting only introduces 4~5bp changes
library(Rsamtools)
#   read in BAM file with ATACseqQC wrapper
#     bigFile=T necessary for large BAMs
alignment <- readBamFile(sample.file, asMates=T, bigFile=F)
#   path to save the shifted BAM - you should not use temporary file
shifted.path <- tempfile("shifted", fileext = ".bam")
#   shift BAM file and grab the shifted alignment object
shifted <- shiftGAlignmentsList(alignment, outbam = shifted.path)

# Get TSS-related quality controls
#   load transcripts of your genome
library(TxDb.Dmelanogaster.UCSC.dm6.ensGene)
txs <- transcripts(TxDb.Dmelanogaster.UCSC.dm6.ensGene)
#   TSS enrichment score
tsse <- TSSEscore(shifted, txs)
tsse$TSSEscore
plot(100*(-9:10-.5), tsse$values, type="b", 
     xlab="distance to TSS",
     ylab="aggregate TSS score")
```

`ATACseqQC` also offers nucleosome positioning by splitting the reads into
nucleosome-free, monucleosome, etc.  This might be done by simply considering 
fragment sizes or using a random forest model taking conservation scores 
into account. Check vignette of `ATACseqQC` for more details.

# Counting transposition event: genomic ranges of interest

To perform differential accessibility analysis, we need to first think about 
which genomic ranges to look at. For instance:

1. 

# Counting transposition events

# Characteristics of the transposition count data

# Differential analysis with `DESeq2`
